\sys: Controllable Program Regeneration

### 2025-06-30 docs/paper/draft.tex
Various tweaks
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\promptbox}[1]{\hspace*{-0.0135\columnwidth}\fbox{\parbox[t]{0.95\columnwidth}{\tt{#1}}}}

@@

\section{\sys Implementation}

@@

\subsection{Example}
\label{sec:example}
This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\texttt{input(P, I), graph(P, I, O), signature(P, S), pure(P)}\}$, where
\texttt{input(P, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.
The, \texttt{pure(P)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.

@@

\begin{wrapfigure}[3]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query requests preservation of I/O behavior and translation to Haskell. The
logic engine deduces that extracting I/O traces and the function signature is
sufficient; the knowledge base automatically associates the target language
\texttt{Haskell} with purity.
\sys transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated 20 I/O pairs, \sys produces the following
Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.5\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \ttt{sig(F)} and \ttt{graph(F)} from the original program, where \texttt{sig(F)} is the
signature of the function (including its name, return type, and arguments), and \texttt{graph(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where the goal is to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until it can make no improvements or there is a negligible difference after two consecutive iterations.
After 15.2 seconds, and two iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[5]{r}{.85\columnwidth}
  \begin{minted}{prolog}
db('music.db'). func(f1). func(f2).
trace_sql(p_, I, T) :- trace_sql(p, I, T), graph(p, I, _).
graph(call(f2, call(f1), I), I, O) :- graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\subsection{Overall Design}
\label{sec:design}

@@

\subsection{Property Language and Extraction}
\label{sec:dsl}

@@

\begin{lstlisting}[caption={A set of facts over a program.}, label=lst:terms]{Name}
language(p, javascript).

\end{lstlisting}

\begin{lstlisting}[caption={Logic rules describing relationships between the original and regenerated program.}]{Name}
graph(p_) :- graph(p).
len(p_, N) :- len(p, N_), N = N_ + 1.
\end{lstlisting}

\begin{lstlisting}[caption={A choice rule.}, label=lst:choice]{Name}
{absent(F) : function(F, P)} :- program(P).
\end{lstlisting}

\begin{lstlisting}[caption={An integrity constraint.}, label=lst:constraint]{Name}
:- language(p, X), language(p_, Y), X != Y.
\end{lstlisting}

\begin{lstlisting}[caption={An optimization directive.}, label=lst:optimization]{Name}
#minimize{C : cost(E,C)}.
\end{lstlisting}

@@

\subsection{Synthesis Plan}
\label{sec:synthesis}

@@

\subsection{Property Verification and Feedback Loop}
\label{sec:verification}

>>>>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\promptbox}[1]{%
  \vspace{0.5em}%
  \hspace*{-0.0135\columnwidth}%
  \fbox{%
    \parbox[t]{0.95\columnwidth}{\tt #1}%
  }%
  \vspace{0.5em} % adjust this value as desired
}

@@

\section{Example}
\label{sec:example}

@@

This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\begin{wrapfigure}[3]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}, \ttt{pure(P)}\}$, where
\texttt{input(P, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.
The, \texttt{pure(P)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.

@@

\begin{wrapfigure}[3]{r}{.7\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query requests preservation of I/O behavior and translation to Haskell. The
logic engine deduces that extracting I/O traces and the function signature is
sufficient; the knowledge base automatically associates the target language
\texttt{Haskell} with purity.
\sys transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated 20 I/O pairs, \sys produces the following
Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \ttt{signature(F)} and \ttt{graph(F)} from
the original program, where \texttt{signature(F)} is the
signature of the function (including its name, return type, and arguments), and \texttt{graph(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where the goal is to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until it can make no improvements or there is a negligible difference after two consecutive iterations.
After 15.2 seconds, and two iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[7]{r}{.8\columnwidth}
  \begin{minted}{prolog}
db('music.db'). func(f1). func(f2).
trace_sql(p_, I, T) :- trace_sql(p, I, T),
                       graph(p, I, _).
graph(call(f2, call(f1), I), I, O) :-
                       graph(p, I, O).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database) driven by an input.
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for 24 generated inputs.
Then, after 20.4 seconds and two attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

\section{Overall Design}
\label{sec:design}

@@

\section{Property Language and Extraction}
\label{sec:dsl}

@@

\begin{listing}
\begin{minted}{prolog}
language(p, javascript).
len(p, 100).
\end{minted}
\caption{A set of facts over a program.}
\label{lst:terms}
\end{listing}

\begin{listing}
\begin{minted}{prolog}
graph(p_) :- graph(p).
len(p_, N) :- len(p, N_), N = N_ + 1.
\end{minted}
\caption{Logic rules describing relationships between the original and regenerated program.}
\label{lst:rules}
\end{listing}

\begin{listing}
\begin{minted}{prolog}
{absent(F) : function(F, P)} :- program(P).
\end{minted}
\caption{A choice rule.}
\label{lst:choice}
\end{listing}

\begin{listing}
\begin{minted}{prolog}
:- language(p, X), language(p_, Y), X != Y.
\end{minted}
\caption{An integrity constraint.}
\label{lst:constraint}
\end{listing}

\begin{listing}
\begin{minted}{prolog}
#minimize{C : cost(E,C)}.
\end{minted}
\caption{An optimization directive.}
\label{lst:optimization}
\end{listing}

@@

\section{Synthesis Plan}
\label{sec:synthesis}

@@

\section{Property Verification and Feedback Loop}
\label{sec:verification}
